Geometric models reveal behavioral and neural signatures of how naturalistic experiences are transformed into episodic memories

### 2020-08-18 paper/main.tex
minor abstract updates

\begin{abstract}
The ways in which our experiences unfold over time are, by nature, incalculably complex.  One effective approach to making sense of this complexity is to characterize experiencess as unique \textit{trajectories} through a certain geometric representational space.  In word embedding spaces, for example, each dimension reflects a concept, such that a given coordinate reflects a weighted mixture of those concepts.  We propose a framework for projecting naturalistic experiences into word embedding spaces, such that the conceptual content of each moment of an experience, and how different moments of the experience relate, are reflected by the \textit{shape} of the experience's trajectory.  By projecting memories of those experiences into the same spaces, one may then geometrically compare the shape of the original experience's trajectory to the shape of how it is remembered later.  According to this view, encoding an experience into memory entails geometrically distorting or transforming the original experience's trajectory.  This translates qualitative questions about how we remember naturalistic experiences into quantitative geometric comparisons.  We applied our framework to data collected as participants watched and verbally recounted a television episode while undergoing functional neuroimaging.  We found that the trajectory of each participant's recall reflected the high-level \textit{essence} (i.e., large-scale narrative summaries) of the episode's trajectory, but participants differed markedly in their memories for low-level features (i.e, small-scale details).  We also identified a network of brain structures that were sensitive to the shape of the episode's trajectory through word embedding space, and an overlapping network that predicted, at the time of encoding, how people would distort (transform) the episode's trajectory when they recounted the episode later.  Our work provides insights into how our brains transform ongoing experiences when we encode them into episodic memories, and provides a formal geometric framework for characterizing the complex dynamic content of naturalistic experiences.
\end{abstract}

@@

\paragraph{Lag conditional probability curve (lag-CRP).} The lag-CRP curve~\citep{Kaha96} reflects the probability of recalling a given item after the just-recalled item, as a function of their relative encoding positions (or \textit{lag}).  In other words, a lag of 1 indicates that a recalled item was presented immediately after the previously recalled item, and a lag of -3 indicates that a recalled item came 3 items before the previously recalled item.  For each recall transition (following the first recall), we computed the lag between the current recall event and the next recall event, normalizing by the total number of possible transitions.  This yielded a number-of-participants (17) by number-of-lags (-29 to +29; 61 lags total) matrix. We averaged over the rows of this matrix to obtain a group-averaged lag-CRP curve (Fig.~\ref{fig:list-learning}B).

>>>>

\begin{abstract}
The ways in which our experiences unfold over time are, by nature, incalculably complex.  One popular approach to making sense of this complexity entails characterizing experiences as unique \textit{trajectories} through a particular geometric representational space whose dimensions are features of our experiences.  In word embedding spaces, for example, each dimension reflects a concept.  And by extension, any given coordinate reflects a weighted mixture of those concepts.  We propose a framework for projecting naturalistic experiences into word embedding spaces, such that the conceptual content of each moment of an experience, and how different moments of the experience relate, are captured by the \textit{shape} of the experience's trajectory.  By projecting memories of those experiences into the same spaces, one may then geometrically compare the shape of the original experience's trajectory to the shape of how it is remembered later.  According to this view, encoding an experience into memory entails geometrically distorting or transforming the original experience's trajectory.  This translates qualitative questions about how we remember naturalistic experiences into quantitative geometric comparisons.  We applied our framework to data collected as participants watched and verbally recounted a television episode while undergoing functional neuroimaging.  We found that the trajectory of each participant's recall reflected the high-level \textit{essence} (i.e., large-scale narrative summaries) of the episode's trajectory, but participants differed markedly in their memories for low-level features (i.e, small-scale details).  We also identified a network of brain structures that were sensitive to the shape of the episode's trajectory through word embedding space, and an overlapping network that predicted, at the time of encoding, how people would distort (transform) the episode's trajectory when they recounted the episode later.  Our work provides insights into how our brains transform ongoing experiences when we encode them into episodic memories, and provides a formal geometric framework for characterizing the complex dynamic content of naturalistic experiences.
\end{abstract}

@@

\paragraph{Lag conditional probability curve (lag-CRP).} The lag-CRP curve~\citep{Kaha96} reflects the probability of recalling a given item after the just-recalled item, as a function of their relative encoding positions (or \textit{lag}).  In other words, a lag of 1 indicates that a recalled item was presented immediately after the previously recalled item, and a lag of -3 indicates that a recalled item came 3 items before the previously recalled item.  For each recall transition (following the first recall), we computed the lag between the current recall event and the next recall event, normalizing by the total number of possible transitions.  This yielded a number-of-participants (17) by number-of-lags (-29 to +29; 59 lags total) matrix. We averaged over the rows of this matrix to obtain a group-averaged lag-CRP curve (Fig.~\ref{fig:list-learning}B).
